Spatial and temporal variability in economic availability of residual biomass from timber harvest in west-central Alberta, Canada (1990-2015)

### 2021-10-17 analysis/manuscript.Rmd
writing at full speed

---
title: Historical Supply of Biomass from Timber Harvest Residue in West-central Alberta, Canada
author:
  - Glen W. Armstrong:
      institute:
        - rr
      correspondence: "yes"
      email: gwa@ualberta.ca
  - Jay A. Anderson:
      email: jaya@ualberta.ca
      institute:
        - rees
  - Alex W. McPhee:
      email: awmcphee@ualberta.ca
      institute: 
        - rees
institute:
  - rr : Department of Renewable Resources, University of Alberta
  - rees: Department of Resource Economics and Environmental Sociology, University of Alberta
date: "`r format(Sys.time(), '%Y-%m-%d')`"
#date: "2021-10-05"
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
output:
  html_document:
    number_sections: yes
    df_print: paged
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  pdf_document:
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  word_document:
     pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
#csl: canadian-journal-of-forest-research
csl: biomass-and-bioenergy
bibliography: historicalResidueSupply.bib
---

<!-- \fontfamily{cmr} -->
<!-- \fontsize{12}{22} -->
<!-- \fontseries{b} -->
<!-- \selectfont -->

@@

Here's something that might help our writing.  It's got the headings that
*Biomass and Bioenergy* wants and presents the main results based on our
placeholder costs.  Author priority is not fixed.

@@

<!-- @R2021 -->

<!-- @Wickham2019 -->

<!-- @qgis2021 -->

@@

@@

@@

#truck_rate <- 6.43 # $/odt/hr
prepiling_cost <- 6.30 # $/odt
comminution_cost <- 25.60 # $/odt
other_cost <- 8.48 # $/odt
constant_cost <- prepiling_cost +  comminution_cost + other_cost # $/Mg DM
location <- c('Whitecourt','Edson','Hinton')
capacity <- c(200000,400000,800000)

merch_vol_adj_factor <- 2.38
odmt_adj_factor <-  0.10

@@

```
# Materials and methods

@@

## Hypothetical plant locations and capacities
There are a number of companies with Forest Management Agreement Areas
located in or near Edson, Whitecourt, and Hinton.
The West Fraser Mills Ltd. (Hinton) FMA is the primary timbershed for the
Hinton Wood Products sawmill and the Hinton Pulp pulp mill.  The West Fraser
Mills Ltd (Edson) FMA is the primary timbershed for the Edson Forest Products 
sawmill.  The Weyerhaeuser Company Limited (Pembina Timberlands) FMA supplies 
the Edson oriented strand board mill and the Drayton Valley sawmill.  The 
Millar Western FMA supplies sawmills in Whitecourt and Fox Creek, a pulp mill
in Whitecourt, and a specialty wood products facility and Acheson.  The Blue
Ridge Lumber Inc. FMA supplies a sawmill in Blue Ridge, which is about 23 km 
from Whitecourt.  The ANC Timber Ltd. FMA supplies logs to several sawmills in
the area, in exchange for chips used to feed the Alberta Newsprint
Company newsprint mill in Whitecourt.

@@

As well as considering locations for the hypothetical plants, we considered 3
different plant capacities: 200,000 Mg DM/yr corresponding to a small wood pellet
mill, 400,000 Mg DM/yr corresponding to a large pellet mill or a small ethanol
plant, and 800,000 Mg DM/yr corresponding to a large ethanol plant.

@@

| Road class      | loaded | unloaded | harmonic mean | 
|:----------------|-------:|---------:|--------------:|
| primary highways|    80  |       95 |           85  |
| other paved     |    65  |       80 |           70  |
| 2-lane gravel   |    50  |       70 |           60  | 
| 1-lane gravel or dirt  |    20  |       30 |           25  | 
| trail           |     ?  |       ?  |           ?   |

@@

# Theory/Calculation
The GIS analysis for this paper was done using QGIS @QGISorg2021.  Data
summaries and statistical analysis were done using R @R2021 using RStudio
@RStudioTeam2021, and tidyverse @Wickham2019.

## Residue yield (Mg DM/ha) by township.

1. Filter out unmerchantable cells from 250 m data
   1. age < 80 years
   2. merchantable volume < 50 m^3^/ha based on 13+/7 cm utilization standard
   8. Bw + Lt prop > 50
2. average merchantable volume of remaining cells to the township level
3. adjust average inventory volume to harvested volume
   - average volume/ha harvested from Hinton FMA / average volume/ha in merchantable inventory from Hinton FMA = 2.38

@@

Following @Roser2013, the pre-piling cost was set to
`r sprintf("%5.2f",prepiling_cost)` \$/Mg DM, the comminution (*grinding ?*) cost
was set to `r sprintf("%5.2f",comminution_cost)`  \$/Mg DM, and other costs were
set to  `r sprintf("%5.2f",other_cost)` \$/Mg DM.  The total "constant" costs are
`r sprintf("%5.2f",constant_cost)` \$/Mg DM.
Using *blah blah blah* , the truck rate was set to 
`r sprintf("%5.2f",truck_rate)` \$/Mg DM/h.
A 100 m³ trailer can haul `r odt_trailer` Mg DM of chipped residue.

```{r read_data_files, echo=FALSE}
t <- read.csv(file = '../data/township_merchantable_volume.csv')
twp_merch_vol <- t %>% select(c(TRM,mean_merch_volume)) 
twp_merch_vol$adj_vol <- twp_merch_vol$mean_merch_volume * merch_vol_adj_factor
#twp_merch_vol$residue_odmt <- twp_merch_vol$adj_vol * odmt_adj_factor
load("../data/biomass_by_trm.RData")
twp_merch_vol <- merge(twp_merch_vol,biomass_by_trm,by = "TRM", all = TRUE)
twp_merch_vol <- twp_merch_vol %>% 
  mutate(adj_factor = res_total/adj_vol)

@@

t <- read.csv(file = '../data/harvest_township_stats_melt_area.csv')
t <- subset(t, select = c(TRM,year,harvest_ha)) %>% filter(year >=1990)
t <- merge(t, twp_merch_vol, by = "TRM", all=TRUE)
t$adj_vol <- t$adj_vol %>% replace_na(0)  # there is an NAs problem here.  replace NAs with average.
                                              # is that the right way to do it?
t$harvest_vol <- t$harvest_ha * t$adj_vol 
t$residue_odmt <- t$harvest_ha * t$res_total
t <- t %>% 
  filter(!is.na(residue_odmt))

@@

#harv2 <- aggregate(harv$harvest_vol,by=list(Category=harv$year), FUN=sum)
harv2 <- harv %>% group_by(year) %>%
  summarise(harvest_vol = sum(harvest_vol)) 
# %>%
#   mutate(hvol2=rollapply(harvest_vol,5,mean,align='left',fill=NA))

#summary(harv2$harvest_vol)
#harv2

harv3 <- read.csv(file = '../data/ABHarvestVol.csv') %>% filter(year >=1990)
#harv3

harv4 <- merge(harv2, harv3, by = "year", all=TRUE)

# ggplot(data=harv4, aes(x=year)) +
#   geom_line(aes(y = harvest_vol), colour="darkred") +
#   geom_line(aes(y = harvvol), colour="steelblue") +
#   labs(x = "year") + labs(y = "harvest volume (million m3)")  +
#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) 
# 
# #2009 is the median harvest volume year for Alberta in 1990-2015
# 
# ggplot(data=harv2, aes(x=year,y=harvest_vol)) +
#   labs(x = "year") +
#   labs(y = "harvest volume (million m3)") +
#   geom_line()  +
#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) 
# +
#   scale_colour_viridis(discrete = TRUE) +
#   theme(legend.position = "none") +
#   facet_grid(cols=vars(location)) +
#   scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))

@@

```
<!-- As a check, we obtained harvest levels for the period 1990-2018 from the National -->
<!-- Forestry Database @ccfm2020 .  Note that in 1998, industrial salvage volumes of 369,791 m³ of  -->
<!-- softwood and 264,647 m³ are not included in the harvest levels shown in the graph -->

<!-- ```{r pressure, echo=FALSE, fig.cap="NFDP harvest levels", out.width = '100%'} -->
<!-- knitr::include_graphics("../img/alberta_harvest.png") -->
<!-- ``` -->

@@

p <- ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (1000 Mg DM)") +
  labs(y = "cycle time (hr)") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))+
  scale_y_continuous(limits=c(0,12))

@@

```{r more_supply_curves, echo=FALSE, warning=FALSE, fig.cap="Residue supply (delivered cost) curves by location.  Each curve represents one year.  The red curve is for 2009, which was the year of median harvest volume."}
ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (1000 Mg DM)") +
  labs(y = "marginal cost ($/Mg DM)") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000)) +
  scale_y_continuous(limits=c(50,175)) 
```

@@

p <- ggplot(t,aes(factor(year),residueCost))  + 
  geom_tufteboxplot() +  
  facet_grid(cols=vars(location),rows=vars(capacity)) +
  labs(y = "residue cost ($/Mg DM)", x="year") +
  scale_x_discrete(breaks=c("1990","","","","","1995","","","","","2000","","","","","2005","","","","","2010","","","","","2015")) +
  theme(axis.text.x = element_text(angle=90))
p

@@

ggplot(data=avet, aes(x=year,y=avecost,colour=location))+
  labs(x = "year") +
  labs(y = "average cost ($/Mg DM)") +
  geom_line() +
  facet_grid(rows=vars(capacity)) +
  scale_y_continuous(limits=c(60,150))

@@

Whitecourt is low cost for most years at all capacities.  Noticable spike in
1999.  Probably due to Virginia Hills fire salvage.  Note that this spike
also affects Hinton and Edson at 800,000 Mg DM capacity, because they reach into the 
Whitecourt area.

@@

This research was supported by funding from the Canada First Research Excellence Fund as part of the University of Alberta’s Future Energy Systems research initiative.

@@

>>>>

---
title: Spatial and temporal variability in economic availability of residual biomass from timber harvest in west-central Alberta, Canada (1990-2015)
author:
  - Glen W. Armstrong:
      institute:
        - rr
      correspondence: "yes"
      email: gwa@ualberta.ca
  - Jay A. Anderson:
      email: jaya@ualberta.ca
      institute:
        - rees
  - Alex W. McPhee:
      email: awmcphee@ualberta.ca
      institute: 
        - rees
institute:
  - rr : Department of Renewable Resources, University of Alberta
  - rees: Department of Resource Economics and Environmental Sociology, University of Alberta
date: "`r format(Sys.time(), '%Y-%m-%d')`"
#date: "2021-10-05"
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
output:
  pdf_document:
    includes:
      in_header: "preamble.tex"
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  html_document:
    number_sections: yes
    df_print: paged
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  word_document:
     pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
#csl: canadian-journal-of-forest-research
csl: biomass-and-bioenergy
bibliography: historicalResidueSupply.bib
---
\raggedright
<!-- \fontfamily{cmr} -->
<!-- \fontsize{12}{22} -->
<!-- \fontseries{b} -->
<!-- \selectfont -->

@@

1. **Focus on cost variability associated with transportation and forest  characteristics**
2. **Based on historical harvest locations (1990-2015)**

@@

@@

```{r BRIMS, echo=FALSE, fig.cap="BRIMS residue", out.width = '100%'}
knitr::include_graphics("../img/BRIMS.png")
```

@@

# Materials and methods

@@

#truck_rate <- 6.43 # $/odt/hr
prepiling_cost <- 6.30 # $/odt
comminution_cost <- 25.60 # $/odt
other_cost <- 8.48 # $/odt
constant_cost <- prepiling_cost +  comminution_cost + other_cost # $·Mg^-1^ (DM)
location <- c('Whitecourt','Edson','Hinton')
capacity <- c(200000,400000,800000)

merch_vol_adj_factor <- 2.77 # based on a wild ass guess
#merch_vol_adj_factor <- 2.38 #based on West Fraser Hinton
odmt_adj_factor <-  0.10

@@

@@

## Hypothetical plant locations and capacities
There are a number of companies with Forest Management Agreement Areas
located near Edson, Whitecourt, and Hinton.
The West Fraser Mills Ltd. (Hinton) FMA is the primary timbershed for the
Hinton Wood Products sawmill and the Hinton Pulp pulp mill.  The West Fraser
Mills Ltd (Edson) FMA is the primary timbershed for the Edson Forest Products 
sawmill.  The Weyerhaeuser Company Limited (Pembina Timberlands) FMA supplies 
the Edson oriented strand board mill and the Drayton Valley sawmill.  The 
Millar Western FMA supplies sawmills in Whitecourt and Fox Creek, a pulp mill
in Whitecourt, and a specialty wood products facility and Acheson.  The Blue
Ridge Lumber Inc. FMA supplies a sawmill in Blue Ridge, which is about 23 km 
from Whitecourt.  The ANC Timber Ltd. FMA supplies logs to several sawmills in
the area, in exchange for chips used to feed the Alberta Newsprint
Company newsprint mill in Whitecourt.

@@

As well as considering locations for the hypothetical plants, we considered 3
different plant capacities: 200 000 Mg/a (DM) corresponding to a small wood pellet
mill, 400 000 Mg/a (DM) corresponding to a large pellet mill or a small ethanol
plant, and 800 000 Mg/a (DM) corresponding to a large ethanol plant.

```{r bb, echo=FALSE, fig.cap="Boreal Bioenergy--Obed.", out.width = '80%'}
knitr::include_graphics("../img/BorealBioenergy.png")
```

@@

| Road class      | loaded | unloaded | harmonic mean | 
|:----------------|-------:|---------:|--------------:|
| primary highways|    80  |       95 |           85  |
| other paved     |    65  |       80 |           70  |
| 2-lane gravel   |    50  |       70 |           60  | 
| 1-lane gravel or dirt  |    20  |       30 |           25  | 
| no road           |     2  |       2  |           2   |

@@

# Theory/Calculation

This is a second best analysis.  Detailed information on harvest volumes
tied to spatial location is held by the provincial government as scale records.
Real (as opposed to modelled) forest inventory information is collected by
holders of forest management agreements as Alberta Vegetation Inventory.
Despite the fact that forest harvesting occurs mostly on public land, the
government considers this information to be the property of companies, and 
requires the consent of companies to release it to researchers.  There are
currently 21 Forest Management Agreements in Alberta.  Permission would need 
to be obtained from the holders of each of those FMAs as well as smaller
companies that operate sawmills.

The GIS analysis for this paper was done using QGIS @QGISorg2021.  Data
summaries and statistical analysis were done using R @R2021 with RStudio
@RStudioTeam2021, and tidyverse @Wickham2019.

## Residue yield (Mg · ha^-1^ (DM)) by township.

### Merchantable volume township

1. Filter out unmerchantable cells from 250 m data
   1. age < 80 years
   2. merchantable volume < 50 m^3^/ha based on 13+/7 cm utilization standard
   8. Bw + Lt prop > 50
2. average merchantable volume of remaining cells to the township level
3. adjust average inventory volume to actual harvested volume from the 
national forestry database @CCFM2021.
   - average volume/ha harvested / inventory volume = 2.77

### Tree characteristics

Most of he forest management plans prepared by FMA holders include projections
of average harvested pieces size, for both needleleafed and broadleafed trees,
if relevant.  We used the piece size reported for the
first 5-year period of the 200-year planning horizon.

We used the harvest piece sizes reported in the detailed forest management
plans for the townships in each FMAA.  For those areas where piece size was
not reported we assumed an average piece size of 0.278 m^3^ for needleleafed
trees and 0.264 m^3^ for broadleafed trees.  These are the average of the values
reported in the forest management plans.  These values are based on the 
utilization
standards used on the FMA.

Utilization standards vary by FMA, and are specified as combinations of 
stump height, stump diameter (outside bark), and minimum log length to a 
specified top diameter.  Stump heights used include 15 cm and 30 cm; stump 
diameters include 15 and 13 cm; top diameters include 12, 11, 10, and 7 cm;
minimum log lengths include 2.44, 2.67, 3.66, 3.76, and 4.88 m.  For areas 
outside of FMAAs we use 30 cm stump height, 15 cm minumum stump diameter, 11
cm top for needleleafed trees and 10 cm top for broadleafed trees, and a 4.88
m minimum log length.  These correspond to the default utilization standards
specified in the Alberta timber harvest planning and operating ground rules
@Alberta2016.

Every township has a leading needleleafed and broadleafed species and
a proportion of merchantable volume associated 
with it derived from the 250 m resolution national inventory @Beaudoin2017.

Ung *et al.* @Ung2013 present taper models and the associated coefficients
for the tree species found in Canada.  These taper models can be used to
find the diameter at any point along the main bole of a tree given species
and diameter at breast height (DBH).  DBH is a commonly used tree measurement
and is the diameter of the tree measured at 1.3 m above ground level, in
most jurisdictions.  It is straightforward to determine cross-sectional area
of the tree at any height.  Integrating this area function will yield volume
between any two heights.

\[
d(d_b,h) = \left(d^2_{b} \frac{\beta_0 d_{b}^{\beta_1}}{\beta_0 d_{b}^{\beta_1} - 1.3} \left(\frac{h}{1.3}\right)^{2-\beta_2}\right)^{-2}
\]

where $d$ is the diameter (cm) at height $h$ (m), and $d_b$ is the diameter at
breast height (1.3 m above ground level)

Implicit in this equation is an equation for tree height

\[
H(d_b) = \beta_0 d_b ^ {\beta_1}
\]

The cross-sectional area equation

\[
a(d_b,h) = \pi \left(\frac{d(d_b,h)}{200}\right)^2
\]

\[
v(d_b,h_s,h_t) = \int_{h_s}^{h_t}\pi \left(\frac{d(d_b,h)}{200}\right)^2dh
\]

The constant 200 is used to convert a diameter (cm) to a radius (m).  
**find a way to reduce possible confusion with $d$**

There is also a simple linear model relating outside bark to inside bark diameter.  The
taper function is for outside bark diameter, but piece sizes are reported
as inside bark volume.  Binary search is used to find the height associated
with the inside bark diameter specified in the utilization standard.

Every combination of township and FMU has a piece size for needleleafed and
broadleafed associated with it, and the leading needleleafed and broadleafed 
species.  We used a binary search procedure to find dbh and height of the 
"average" broadleafed and needleleafed tree in each township based on the 
companies utilization standard.  An average piece size to the 13+/7 utilization
standard was also calculated by integrating the area function using the 
appropriate limits.

Using these equations, we are able to calculate wood volumes and bark volumes
between any two heights along the tree stem.  We calculate the proportions 
of total tree and bark volumes in the top of the tree.  We use this volume 
proportion later to calculate biomass of stemwood and bark in the top.

The number of harvested stems per ha is calculated by dividing the 13/7 volume
per ha by the 13/7 piece size.

### Individual tree biomass.

Canada's national forest inventory website provides a useful individual
tree biomass calculator @NFIS2016.  It calculates the biomass of tree 
components (stem wood, stem bark, branches, foliage, and total) using
province/territory, terrestrial ecozon, species, dbh, and tree height as input.
It is possible to upload a file to the calculator so that the biomass of
many trees can be calculated at once.  The calculator is based on work
by Lambert *et al.* @Lambert2005.

We assume that the residual biomass is in the stem wood and stem bark from the
tops of the harvested trees and the branches and foliage.  The residual biomass
from the tops is calculated using the total biomass of stem wood and stem bark
adjusted by the volume proportions calculated above.

### Township level biomass

For both needleleaf and broadleaf, we multiply the residue associated with the
individual tree by the number of harvested stems per ha.  We reduce this
value using a recovery factor of 0.645 **??** following Peltola *et al.* 
@Peltola2011, recognizing that some of the residual biomass will 
be unrecoverable.

```{r resdiue, echo=FALSE, fig.cap="Biomass residue available from harvest.", out.width = '100%'}
knitr::include_graphics("../img/residue_per_ha.png")
```

@@

```{r roser, echo=FALSE, fig.cap="Röser.", out.width = '100%'}
knitr::include_graphics("../img/roser.png")
```

```{r hourly_rates, echo=FALSE, fig.cap="Hourly rates for forestry equipment.", out.width = '100%'}
knitr::include_graphics("../img/hourly_rates.png")
```

Following @Roser2013, the pre-piling cost was set to
`r sprintf("%5.2f",prepiling_cost)` \$·Mg^-1^ (DM), the comminution (*grinding ?*) cost
was set to `r sprintf("%5.2f",comminution_cost)`  \$·Mg^-1^ (DM), and other costs were
set to  `r sprintf("%5.2f",other_cost)` \$·Mg^-1^ (DM).  The total "constant" costs are
`r sprintf("%5.2f",constant_cost)` \$·Mg^-1^ (DM).
Using *blah blah blah* , the truck rate was set to 
`r sprintf("%5.2f",truck_rate)` \$·Mg^-1^ (DM)/h.
A 100 m³ trailer can haul `r odt_trailer` Mg (DM) of chipped residue.

```{r read_data_files, echo=FALSE}
newAB <- read.csv(file = '../data/AlbertaHarvestLevels.csv',
                  fileEncoding = 'UTF-8-BOM')
t <- read.csv(file = '../data/township_merchantable_volume.csv')
twp_merch_vol <- t %>% select(c(TRM,mean_merch_volume)) 
twp_merch_vol$adj_vol <- twp_merch_vol$mean_merch_volume * merch_vol_adj_factor
#twp_merch_vol$residue_odmt <- twp_merch_vol$adj_vol * odmt_adj_factor
load("../data/biomass_by_trm.RData")
twp_merch_vol_temp <- merge(twp_merch_vol,biomass_by_trm,by = "TRM", all = TRUE)
twp_merch_vol <- twp_merch_vol_temp %>% 
  mutate(adj_factor = res_total/adj_vol)

@@

t <- read.csv(file = '../data/harvest_township_stats_melt_area.csv')
t <- subset(t, select = c(TRM,year,harvest_ha)) %>% filter(year >=1990)
t <- merge(t, twp_merch_vol, by = "TRM", all=TRUE)
t$adj_vol <- t$adj_vol %>% replace_na(0)  # there is an NAs problem here.  replace NAs with average.
                                              # is that the right way to do it?
t$unadj_harvest_vol <-t$harvest_ha * t$mean_merch_volume
t$harvest_vol <- t$harvest_ha * t$adj_vol 
t$residue_odmt <- t$harvest_ha * t$res_total
t <- t %>% 
  filter(!is.na(residue_odmt))

@@

#harv2 <- aggregate(harv$harvest_vol,by=list(Category=harv$year), FUN=sum)
harv2 <- harv %>% group_by(year) %>%
  summarise(harvest_vol = sum(harvest_vol),
            unadj_harvest_vol = sum(unadj_harvest_vol))

# %>%
#   mutate(hvol2=rollapply(harvest_vol,5,mean,align='left',fill=NA))

#summary(harv2$harvest_vol)
#harv2
# ggplot(data=harv2, aes(x=year)) +
#   geom_line(aes(y = harvest_vol), colour="darkred") +
# #  geom_line(aes(y = harvvol), colour="steelblue") +
#   labs(x = "year") + labs(y = "harvest volume (million m3)")  +
#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6))

#2009 is the median harvest volume year for Alberta in 1990-2015

# ggplot(data=harv2, aes(x=year,y=harvest_vol)) +
#   labs(x = "year") +
#   labs(y = "harvest volume (million m3)") +
#   geom_line()  +
#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6))
# +
#   scale_colour_viridis(discrete = TRUE) +
#   theme(legend.position = "none") +
#   facet_grid(cols=vars(location)) +
#   scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))
harv3 <- read.csv(file = '../data/ABHarvestVol.csv') %>% filter(year >=1990)
#harv3

harv4 <- merge(harv2, harv3, by = "year", all=TRUE)
harv5 <- merge(harv4, newAB, by = "year", all=TRUE)

harv_comparison <- ggplot(data=harv5, aes(x=year,y=harvest_vol, colour="Estimated harvest")) +
  labs(x = "year") +
  labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
  geom_line()  +
  geom_line(aes(y=total, colour="Actual harvest"))  +
  scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6))
# +
#   scale_colour_viridis(discrete = TRUE) +
#   theme(legend.position = "none") +
#   facet_grid(cols=vars(location)) +
#   scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))
# 
# head(harv4)

@@

As a check, we obtained harvest volumes from Alberta for the period 1990-2015 
from the national
forestry database @CCFM2021.  
Note that in 1998, industrial salvage volumes of 369 791 m^3^ of
softwood and 264 647 m^3^ are not included in the harvest levels 
shown in the graph.

```{r pressure, echo=FALSE, fig.cap="NFDP harvest levels compared with estimated.", out.width = '100%'}
#knitr::include_graphics("../img/alberta_harvest.png")
harv_comparison
```

```{r harvbyowner, echo=FALSE, fig.cap="NFDP harvest levels by ownership.", out.width = '100%'}
knitr::include_graphics("../img/harvestbyowner.png")
```

@@

p <- ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (1000 Mg (DM))") +
  labs(y = "cycle time (hr)") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))+
  scale_y_continuous(limits=c(0,12))

@@

```{r more_supply_curves, echo=FALSE, warning=FALSE, fig.cap="Residue supply (delivered cost) curves by location.  Each curve represents one year.  The red curve is for 2009, which was the year of median harvest volume."}
ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (1000 Mg (DM))") +
  labs(y = "marginal cost ($·Mg^-1^ (DM))") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000)) +
  scale_y_continuous(limits=c(50,175)) 
```

@@

p <- ggplot(t,aes(factor(year),residueCost))  + 
  geom_tufteboxplot() +  
  facet_grid(cols=vars(location),rows=vars(capacity)) +
  labs(y = "residue cost ($·Mg^-1^ (DM))", x="year") +
  scale_x_discrete(breaks=c("1990","","","","","1995","","","","","2000","","","","","2005","","","","","2010","","","","","2015")) +
  theme(axis.text.x = element_text(angle=90))
p

@@

ggplot(data=avet, aes(x=year,y=avecost,colour=location))+
  labs(x = "year") +
  labs(y = "average cost ($·Mg^-1^ (DM))") +
  geom_line() +
  facet_grid(rows=vars(capacity)) +
  scale_y_continuous(limits=c(60,150))

@@

Whitecourt is low cost for most years at all capacities.  Noticable spike in
1999.  Probably due to Virginia Hills fire salvage.  Note that this spike
also affects Hinton and Edson at 800 000 Mg/a (DM) capacity, 
because they reach into the 
Whitecourt area.

@@

This research was supported by funding from the Canada First Research Excellence
Fund as part of the University of Alberta’s Future Energy Systems 
research initiative
<!-- (Project # CFREF-2015-00001) -->
.

@@

We work with a group of talented people.  Thanks to Marty Luckert, Feng Qiu,
Grant Hauer, and Wenbei Zhang for helpful discussion.
